Generating Random Logic Programs Using Constraint Programming

### 2020-05-21 text/paper2/paper.tex
Incorporating some feedback

\documentclass[runningheads]{llncs}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}
\usepackage{mathrsfs}
\usepackage[nounderscore]{syntax}
\usepackage{blkarray}
\usepackage[binary-units]{siunitx}
\usepackage[inline]{enumitem}
\usepackage{capt-of}
\usepackage[caption=false]{subfig}
\usepackage{booktabs}

@@

%The abstract should briefly summarize the contents of the paper in
%150--250 words.
\begin{abstract}
  We present a novel approach to generating random logic programs and
  random probabilistic logic programs using constraint programming. The
  generated programs are useful in empirical testing of inference algorithms,
  random data generation, and program learning. This approach has a major
  advantage in that one can easily add additional conditions for the generated
  programs. As an example of this, we introduce a new constraint for predicate
  independence with efficient propagation and entailment algorithms, allowing
  one to generate programs that have a certain independence structure. Finally,
  we provide a combinatorial argument for the correctness of the model,
  describe how the parameters of the model affect the empirical difficulty of
  the program generation task, and use the model to compare probabilistic
  inference algorithms across a range of synthetic problems.

@@

% TODO: proofread, run a grammar checker on everything

@@

Unifying logic and probability is a long-standing challenge in artificial
intelligence \cite{DBLP:journals/cacm/Russell15}, and, in that regard,
statistical relational learning (SRL) has developed into an exciting area that
mixes machine learning and symbolic (logical and relational) structures. In
particular, probabilistic logic programs---including languages such as
\textsc{PRISM} \cite{DBLP:conf/ijcai/SatoK97}, \textsc{ICL}
\cite{DBLP:journals/ai/Poole97}, and \textsc{ProbLog}
\cite{DBLP:conf/ijcai/RaedtKT07}---are promising frameworks for codifying
complex SRL models. With the enhanced structure, however, inference becomes more
challenging as algorithms have to correctly handle hard and soft logical
constraints. At the moment, we have no precise way of evaluating and comparing
different inference algorithms. Incidentally, if one were to survey the
literature, one often finds that an inference algorithm is only tested on 1--4
data sets
\cite{DBLP:conf/ecai/BruynoogheMKGVJR10,DBLP:journals/tplp/KimmigDRCR11,DBLP:conf/ijcai/VlasselaerBKMR15},
originating from areas such as social networks, citation patterns, and
biological data. But how confidently can we claim that an algorithm works well
if it is only tested on a few types of problems?

@@

In \cref{sec:heads,sec:bodies,sec:variable_symmetry}, we introduce a
constraint-based representation for logic programs based on simple parameters
that describe the program's size, what predicates and constants it uses, etc.
This representation takes the form of a \emph{constraint satisfaction problem}
(CSP), i.e., a set of discrete variables and restrictions on what values they
can take. Every solution to this problem (as output by a constraint solver)
directly translates into a logic program. One can either output all
(sufficiently small) programs that satisfy the given conditions or use random
value-ordering heuristics and restarts to generate random programs. For sampling
from a uniform distribution, the CSP can be transformed into a belief network
\cite{DBLP:conf/aaai/DechterKBE02}. In fact, the same model can generate both
probabilistic programs in the syntax of \textsc{ProbLog}
\cite{DBLP:conf/ijcai/RaedtKT07} and non-probabilistic \textsc{Prolog} programs.
This is both the first time the problem of generating random logic programs has
been approached in its full generality (i.e., including first-order clauses with
variables) and the first time inference algorithms for probabilistic logic
programs can be evaluated on more than a handful of instances.

@@

The basic primitives of logic programs are \emph{constants}, \emph{(logic)
  variables}, and \emph{predicates}. Each predicate has an \emph{arity} that
defines the number of terms that it can be applied to. A \emph{term} is either a
variable or a constant, and an \emph{atom} is a predicate of arity $n$ applied
to $n$ terms. A \emph{formula} is any well-formed expression that connects
atoms using conjunction $\land$, disjunction $\lor$, and negation
$\neg$. A \emph{clause} is a pair of a \emph{head} (which is an atom) and a
\emph{body} (which is a formula\footnote{Our model supports arbitrarily complex
  bodies of clauses because \textsc{ProbLog} does too. However, one can easily
  restrict our representation of a body to a single disjunction of literals by
  adding a couple of additional constraints.}). A \emph{(logic) program} is a
set of clauses, and a \emph{\textsc{ProbLog} program} is a set of
clause-probability pairs.

@@

We need to constrain the forest represented by the $\variable{structure}$
array together with its $\variable{values}$ to eliminate symmetries and adhere
to our desired format. First, we can recognise that the order of the elements in
the $\variable{structure}$ array does not matter, i.e., the structure is only
defined by how the elements link to each other, so we can add a constraint
saying that $\variable{structure}$ is sorted. Next, since we already have a
variable that counts the number of nodes in the main tree, we can fix the
structure and the values of the remaining trees to some constant values.
\begin{constraint}
  For $i = 1, \dots, \maxNumNodes{} - 1$, if $i < \variable{numNodes}$, then
  \[
    \variable{structure}[i] = i, \quad \text{and} \quad
    \variable{values}[i].\variable{name} = \top,
  \]
  else $\variable{structure}[i] < i$.
\end{constraint}
The second part of this constraint states that every node in the main tree
except the zeroth node cannot be a root and must have its parent located to
the left of itself. Next, we classify all nodes into three classes: predicate
(or empty) nodes, negation nodes, and conjunction/disjunction nodes based on the
number of children (zero, one, and two, respectively).
\begin{constraint} \label{constraint:node_types}
  For $i = 0, \dots, \maxNumNodes{} - 1$, let $C_i$ be the number of times $i$
  appears in the \variable{structure} array with index greater than $i$. Then
  \begin{align*}
    C_i = 0 &\iff \variable{values}[i].\variable{name} \in \predicates{} \cup \{ \top \},\\
    C_i = 1 &\iff \variable{values}[i].\variable{name} = \neg,\\
    C_i > 1 &\iff \variable{values}[i].\variable{name} \in \{ \land, \lor \}.
  \end{align*}
\end{constraint}
The value $\top$ serves a twofold purpose: it is used as the fixed value for
nodes outside the main tree, and, when located at the zeroth node, it can
represent a clause with an empty body. Thus, we can say that only root nodes can
have $\top$ as the value.
\begin{constraint}
  For $i = 0, \dots, \maxNumNodes{} - 1$,
  \[
    \variable{structure}[i] \ne i \implies
    \variable{values}[i].\variable{name} \ne \top.
  \]
\end{constraint}
Finally, we add a way to disable a clause by setting its head predicate to
$\Box$.
\begin{constraint}
  For $i = 0, \dots, \maxNumClauses{} - 1$, if
  $\variable{heads}[i].\variable{predicate} = \Box$, then
  \[
    \variable{bodies}[i].\variable{numNodes} = 1, \quad \text{and}
    \quad \variable{bodies}[i].\variable{values}[0].\variable{name} = \top.
  \]
\end{constraint}

@@

We end the section with several redundant constraints. First, we can state that
the positions occupied by different terms must be different.
\begin{constraint} \label{constraint:all_diff}
  For $u \ne v \in \constants{} \cup \variables{} \cup \{ \Box \}$,
  $\variable{occ}[u] \cap \variable{occ}[v] = \emptyset$.
\end{constraint}
The reason why we use zero to represent an unused variable is so that we could
now use the `all different except zero' constraint for the $\variable{intros}$
array. We can also add another link between $\variable{intros}$ and
$\variable{occ}$ that essentially says that the smallest element of a set is an
element of the set.
\begin{constraint}
  For $v \in \variables{}$, $\variable{intros}[v] \ne 0 \iff
  \variable{intros}[v] - 1 \in \variable{occ}[v]$.
\end{constraint}
Finally, we define an auxiliary set variable to act as a set of possible values
that $\variable{intros}$ can take. Let $\variable{potentials} \subseteq \{ 0,
\dots, N \}$ be such that for $v \in \variables{}$, $\variable{intros}[v] \in
\variable{potentials}$. Using this new variable, we can add a constraint saying
that non-predicate nodes in the tree representation of a clause cannot have
variables as arguments.
\begin{constraint} \label{constraint:potentialIntroductions}
  For $i = 0, \dots, \maxNumNodes{} - 1$, let
  \[
    S = \{ \maxArity{} \times (i + 1) + j + 1 \mid j = 0, \dots, \maxArity{} - 1
    \}.
  \]
  If $\variable{values}[i].\variable{name} \not\in \predicates{}$, then
  $\variable{potentials} \cap S = \emptyset$.
\end{constraint}

@@

To demonstrate the correctness of the model and explain it in more
detail, in this section we are going to derive combinatorial expressions for
counting the number of programs with up to $\maxNumClauses{}$ clauses and up to
$\maxNumNodes{}$ nodes per clause, and arbitrary $\predicates{}$,
$\arities{}$, $\variables{}$, and $\constants{}$\footnote{We checked that
  our model agrees with the derived combinatorial formula in up to a thousand
  different scenarios. The details of this empirical investigation are omitted
  as they are not crucial to the thrust of this paper.}. Being able to establish
two ways to generate the same sequence of numbers (i.e., numbers of programs
with certain properties and parameters) allows us to gain confidence that the
constraint model accurately matches our intentions. For this section, we
introduce the term \emph{total arity} of a body of a clause to refer to the sum
total of arities of all predicates in the body.

@@

Let $\mathscr{P}$ be a probabilistic logic program with its set of predicates
$\predicates{}$. Its \emph{(predicate) dependency graph} is a directed graph
$G_{\mathscr{P}}$ with elements of $\predicates{}$ as nodes and an edge between
$\mathsf{P}, \mathsf{Q} \in \predicates{}$ if there is a clause in $\mathscr{P}$
with $\mathsf{Q}$ as the head and $\mathsf{P}$ mentioned in the body. We say
that the edge is \emph{negative} if there exists a clause with $\mathsf{Q}$ as
the head and at least one instance of $\mathsf{P}$ at the body such that the
path from the root to the $\mathsf{P}$ node in the tree representation of the
clause passes through at least one negation node; otherwise, it is
\emph{positive}. We say that $\mathscr{P}$ (or $G_{\mathscr{P}}$) has a
\emph{negative cycle} if $G_{\mathscr{P}}$ has a cycle with at least one
negative edge. A program $\mathscr{P}$ is \emph{stratified} if $G_{\mathscr{P}}$
has no negative cycles\footnote{This definition is an extension of a well-known
  result for logic programs \cite{DBLP:journals/jlp/BalbinPRM91} to
  probabilistic logic programs with arbitrary complex clause bodies.}. Thus a
simple entailment algorithm for stratification can be constructed by selecting
all clauses, all predicates of which are fully determined, and looking for
negative cycles in the dependency graph constructed based on those clauses using
an algorithm such as Bellman-Ford.

@@

Along with constraints, variables, and their domains, two more design decisions
are needed to complete the model: heuristics and restarts. By trial and error,
the variable ordering heuristic was devised to eliminate sources of
\emph{thrashing}, i.e., situations where a contradiction is being `fixed' by
making changes that have no hope of fixing the contradiction. Thus, we partition
all decision variables into an ordered list of groups and require the values of
all variables from one group to be determined before moving to the next group.
Within each group, we use the `fail first' variable ordering heuristic. The
first group consists of all head predicates. Afterwards, we handle all remaining
decision variables from the first clause before proceeding to the next. The
decision variables within each clause are divided into
\begin{enumerate*}
\item the $\variable{structure}$ array,
\item body predicates,
\item head arguments,
\item (if $|\variables{}| > 1$) the $\variable{intros}$ array,
\item body arguments.
\end{enumerate*}
For instance, in the clause from \cref{example:sibling}, all visible parts of
the clause would be decided in this order:
\[
  \overset{1}{\mathsf{sibling}}(\overset{3}{X}, \overset{3}{Y}) \gets
  \overset{2}{\mathsf{parent}}(\overset{4}{X}, \overset{4}{Z})
  \overset{2}{\land} \overset{2}{\mathsf{parent}}(\overset{4}{Y},
  \overset{4}{Z}).
\]
We also employ a geometric restart policy, restarting after $10, 10 \times 1.1,
10 \times 1.1^2, \dots$ contradictions\footnote{Restarts help overcome early
  mistakes in the search process but can be disabled if one wants to find all
  solutions, in which case search is complete regardless of the variable
  ordering heuristic.}. We ran close to \num{400000} experiments, investigating
whether the model is efficient enough to generate reasonably-sized programs and
gaining insight into what parameter values make the CSP harder. For
$|\predicates{}|$, $|\variables{}|$, $|\constants{}|$, $\maxNumNodes{}$, and
$\maxNumClauses{} - |\predicates{}|$ (i.e., the number of clauses in addition to
the mandatory $|\predicates{}|$ clauses), we assign all combinations of 1, 2, 4,
8. $\maxArity{}$ is assigned to values 1--4. For each $|\predicates{}|$, we also
iterate over all possible numbers of independent pairs of predicates, ranging
from 0 up to $\binom{|\predicates{}|}{2}$. For each combination of the
above-mentioned parameters, we pick ten random ways to assign arities to
predicates (such that $\maxArity{}$ occurs at least once) and ten random
combinations of independent pairs.

@@

For this experiment, we consider clauses of two types: \emph{rules} are clauses
such that the head atom has at least one variable, and \emph{facts} are clauses
with empty bodies and no variables. We use our constraint model to generate the
rules according to the following parameter values: $|\predicates{}|,
|\variables{}|, \maxNumNodes{} \in \{ 2, 4, 8 \}$, $\maxArity{} \in \{ 1, 2, 3
\}$, $\maxNumClauses{} = |\predicates{}|$, $\constants{} = \emptyset$. Just like
before, we explore all possible numbers of independent predicate pairs. We also
add a constraint that forbids empty bodies. For both rules and facts,
probabilities are uniformly sampled from $\{ 0.1, 0.2, \dots, 0.9 \}$.
Furthermore, all rules are probabilistic, while we vary the proportion of
probabilistic facts among \SI{25}{\percent}, \SI{50}{\percent}, and
\SI{75}{\percent}. For generating facts, we consider $|\constants{}| \in \{100,
200, 400 \}$ and vary the number of facts among $10^3$, $10^4$, and $10^5$ but
with one exception: the number of facts is not allowed to exceed
\SI{75}{\percent} of all possible facts with the given values of
$\predicates{}$, $\arities{}$, and $\constants{}$. Facts are generated using a
simple procedure that randomly selects a predicate, combines it with the right
number of constants, and checks whether the generated atom is already included
or not. We randomly select configurations from the description above and
generate ten programs with a complete restart of the constraint solver before
the generation of each program, including choosing different arities and
independent pairs. Finally, we set the query of each program to a random fact
not explicitly included in the program and consider six natively supported
algorithms and knowledge compilation techniques: binary decision diagrams (BDDs)
\cite{DBLP:journals/tc/Bryant86}, negation normal form (NNF), deterministic
decomposable NNF (d-DNNF) \cite{DBLP:journals/jair/DarwicheM02}, K-Best
\cite{DBLP:conf/pkdd/GutmannKKR08}, and two encodings based on sentential
decision diagrams \cite{DBLP:conf/ijcai/Darwiche11}, one of which encodes the
entire program (SDDX), while the other one encodes only the part of the program
relevant to the query (SDD)\footnote{Forward SDDs (FSDDs) and forward BDDs
  (FBDDs)
  \cite{DBLP:journals/corr/abs-1911-07750,DBLP:conf/ijcai/VlasselaerBKMR15} are
  omitted because the former uses too much memory and the implementation of the
  latter seems to be broken at the time of writing.}.

@@

>>>>

\documentclass[runningheads]{llncs}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}
\usepackage{mathrsfs}
\usepackage[nounderscore]{syntax}
\usepackage{blkarray}
\usepackage[binary-units]{siunitx}
\usepackage[inline,shortlabels]{enumitem}
\usepackage{capt-of}
\usepackage[caption=false]{subfig}
\usepackage{booktabs}

@@

\begin{abstract}
  Testing algorithms across a wide range of problem instances is crucial to
  ensure the validity of any claim about one algorithm's superiority over
  another. However, when it comes to inference algorithms for probabilistic
  logic programs, experimental evaluations are limited to only a few programs.
  Existing methods to generate random logic programs are limited to
  propositional programs and often impose stringent syntactic restrictions. We
  present a novel approach to generating random logic programs and random
  probabilistic logic programs using constraint programming, introducing a new
  constraint to control the independence structure of the underlying probability
  distribution. We also provide a combinatorial argument for the correctness of
  the model, show how the model scales with parameter values, and use the model
  to compare probabilistic inference algorithms across a range of synthetic
  problems. Our model allows inference algorithm developers to evaluate and
  compare the algorithms across a wide range of instances, providing a detailed
  picture of their (comparative) strengths and weaknesses.

@@

@@

Unifying logic and probability is a long-standing challenge in artificial
intelligence \cite{DBLP:journals/cacm/Russell15}, and, in that regard,
statistical relational learning (SRL) has developed into an exciting area that
mixes machine learning and symbolic (logical and relational) structures. In
particular, probabilistic logic programs---including languages such as
\textsc{PRISM} \cite{DBLP:conf/ijcai/SatoK97}, \textsc{ICL}
\cite{DBLP:journals/ai/Poole97}, and \textsc{ProbLog}
\cite{DBLP:conf/ijcai/RaedtKT07}---are promising frameworks for codifying
complex SRL models. With the enhanced structure, however, inference becomes more
challenging. At the moment, we have no precise way of evaluating and comparing
inference algorithms. Incidentally, if one were to survey the literature, one
often finds that an inference algorithm is only tested on a small number (1--4)
of data sets
\cite{DBLP:conf/ecai/BruynoogheMKGVJR10,DBLP:journals/tplp/KimmigDRCR11,DBLP:conf/ijcai/VlasselaerBKMR15},
originating from areas such as social networks, citation patterns, and
biological data. But how confident can we be that an algorithm works well if it
is only tested on a few problems?

@@

In this work (\cref{sec:heads,sec:bodies,sec:variable_symmetry}), we introduce a
constraint-based representation for logic programs based on simple parameters
that describe the program's size, what predicates and constants it uses, etc.
This representation takes the form of a \emph{constraint satisfaction problem}
(CSP), i.e., a set of discrete variables and restrictions on what values they
can take. Every solution to this problem (as output by a constraint solver)
directly translates into a logic program. One can either output all
(sufficiently small) programs that satisfy the given conditions or use random
value-ordering heuristics and restarts to generate random programs. For sampling
from a uniform distribution, the CSP can be transformed into a belief network
\cite{DBLP:conf/aaai/DechterKBE02}. In fact, the same model can generate both
probabilistic programs in the syntax of \textsc{ProbLog}
\cite{DBLP:conf/ijcai/RaedtKT07} and non-probabilistic \textsc{Prolog} programs.
To the best of our knowledge, this is the first work that
\begin{enumerate*}[(a)]
\item addresses the problem of generating random logic programs in its full
  generality (i.e., including first-order clauses with variables), and
\item compares and evaluates inference algorithms for probabilistic logic
  programs on more than a handful of instances.
\end{enumerate*}

@@

The basic primitives of logic programs are \emph{constants}, \emph{(logic)
  variables}, and \emph{predicates} with their \emph{arities}. A \emph{term} is
either a variable or a constant, and an \emph{atom} is a predicate of arity $n$
applied to $n$ terms. A \emph{formula} is any well-formed expression that
connects atoms using conjunction $\land$, disjunction $\lor$, and negation
$\neg$. A \emph{clause} is a pair of a \emph{head} (which is an atom) and a
\emph{body} (which is a formula\footnote{Our model supports arbitrarily complex
  bodies of clauses (e.g., $\neg\mathsf{P}(X) \lor (\mathsf{Q}(X) \land
  \mathsf{P}(X))$) because \textsc{ProbLog} does too. However, one can easily
  restrict our representation of a body to a single conjunction of literals
  (e.g., $\mathsf{Q}(X) \land \neg\mathsf{P}(X))$) by adding a couple of
  additional constraints.}). A \emph{(logic) program} is a set of clauses, and a
\emph{\textsc{ProbLog} program} is a set of clause-probability pairs
\cite{DBLP:journals/tplp/FierensBRSGTJR15}.

@@

We need to constrain the forest represented by the $\variable{structure}$
array together with its $\variable{values}$ to eliminate symmetries and adhere
to our desired format. First, we can recognise that the order of the elements in
the $\variable{structure}$ array does not matter, i.e., the structure is only
defined by how the elements link to each other, so we can add a constraint for
sorting the $\variable{structure}$ array. Next, since we already have a
variable that counts the number of nodes in the main tree, we can fix the
structure and the values of the remaining trees to some constant values.
\begin{constraint}
  For $i = 1, \dots, \maxNumNodes{} - 1$, if $i < \variable{numNodes}$, then
  \[
    \variable{structure}[i] = i, \quad \text{and} \quad
    \variable{values}[i].\variable{name} = \top,
  \]
  else $\variable{structure}[i] < i$.
\end{constraint}
The second part of this constraint states that every node in the main tree
except the zeroth node cannot be a root and must have its parent located to
the left of itself. Next, we classify all nodes into three classes: predicate
(or empty) nodes, negation nodes, and conjunction/disjunction nodes based on the
number of children (zero, one, and two, respectively).
\begin{constraint} \label{constraint:node_types}
  For $i = 0, \dots, \maxNumNodes{} - 1$, let $C_i$ be the number of times $i$
  appears in the \variable{structure} array with index greater than $i$. Then
  \begin{align*}
    C_i = 0 &\iff \variable{values}[i].\variable{name} \in \predicates{} \cup \{ \top \},\\
    C_i = 1 &\iff \variable{values}[i].\variable{name} = \neg,\\
    C_i > 1 &\iff \variable{values}[i].\variable{name} \in \{ \land, \lor \}.
  \end{align*}
\end{constraint}
The value $\top$ serves a twofold purpose: it is used as the fixed value for
nodes outside the main tree, and, when located at the zeroth node, it can
represent a clause with an empty body. Thus, we can say that only root nodes can
have $\top$ as the value.
\begin{constraint}
  For $i = 0, \dots, \maxNumNodes{} - 1$,
  \[
    \variable{structure}[i] \ne i \implies
    \variable{values}[i].\variable{name} \ne \top.
  \]
\end{constraint}
Finally, we add a way to disable a clause by setting its head predicate to
$\Box$.
\begin{constraint}
  For $i = 0, \dots, \maxNumClauses{} - 1$, if
  $\variable{heads}[i].\variable{predicate} = \Box$, then
  \[
    \variable{bodies}[i].\variable{numNodes} = 1, \quad \text{and}
    \quad \variable{bodies}[i].\variable{values}[0].\variable{name} = \top.
  \]
\end{constraint}

@@

We end the section with several redundant constraints that make the CSP easier
to solve. First, we can state that the positions occupied by different terms
must be different.
\begin{constraint} \label{constraint:all_diff}
  For $u \ne v \in \constants{} \cup \variables{} \cup \{ \Box \}$,
  $\variable{occ}[u] \cap \variable{occ}[v] = \emptyset$.
\end{constraint}
The reason why we use zero to represent an unused variable is so that we could
now use the `all different except zero' constraint for the $\variable{intros}$
array. We can also add another link between $\variable{intros}$ and
$\variable{occ}$ that essentially says that the smallest element of a set is an
element of the set.
\begin{constraint}
  For $v \in \variables{}$, $\variable{intros}[v] \ne 0 \iff
  \variable{intros}[v] - 1 \in \variable{occ}[v]$.
\end{constraint}
Finally, we define an auxiliary set variable to act as a set of possible values
that $\variable{intros}$ can take. Let $\variable{potentials} \subseteq \{ 0,
\dots, N \}$ be such that for $v \in \variables{}$, $\variable{intros}[v] \in
\variable{potentials}$. Using this new variable, we can add a constraint saying
that non-predicate nodes in the tree representation of a clause cannot have
variables as arguments.
\begin{constraint} \label{constraint:potentialIntroductions}
  For $i = 0, \dots, \maxNumNodes{} - 1$, let
  \[
    S = \{ \maxArity{} \times (i + 1) + j + 1 \mid j = 0, \dots, \maxArity{} - 1
    \}.
  \]
  If $\variable{values}[i].\variable{name} \not\in \predicates{}$, then
  $\variable{potentials} \cap S = \emptyset$.
\end{constraint}

@@

To demonstrate the correctness of the model, in this section we are going to
derive combinatorial expressions for counting the number of programs with up to
$\maxNumClauses{}$ clauses and up to $\maxNumNodes{}$ nodes per clause, and
arbitrary $\predicates{}$, $\arities{}$, $\variables{}$, and $\constants{}$.
Being able to establish two ways to generate the same sequence of numbers (i.e.,
numbers of programs with certain properties and parameters) allows us to gain
confidence that the constraint model accurately matches our intentions. For this
section, we introduce the term \emph{total arity} of a body of a clause to refer
to the sum total of arities of all predicates in the body.

@@

Let $\mathscr{P}$ be a probabilistic logic program with its set of predicates
$\predicates{}$. Its \emph{(predicate) dependency graph} is a directed graph
$G_{\mathscr{P}}$ with elements of $\predicates{}$ as nodes and an edge between
$\mathsf{P}, \mathsf{Q} \in \predicates{}$ if there is a clause in $\mathscr{P}$
with $\mathsf{Q}$ as the head and $\mathsf{P}$ mentioned in the body. We say
that the edge is \emph{negative} if there exists a clause with $\mathsf{Q}$ as
the head and at least one instance of $\mathsf{P}$ at the body such that the
path from the root to the $\mathsf{P}$ node in the tree representation of the
clause passes through at least one negation node; otherwise, it is
\emph{positive}. We say that $\mathscr{P}$ (or $G_{\mathscr{P}}$) has a
\emph{negative cycle} if $G_{\mathscr{P}}$ has a cycle with at least one
negative edge. A program $\mathscr{P}$ is \emph{stratified} if $G_{\mathscr{P}}$
has no negative cycles.\footnote{This definition is an extension of a well-known
  result for logic programs \cite{DBLP:journals/jlp/BalbinPRM91} to
  probabilistic logic programs with arbitrary complex clause bodies.} Thus a
simple entailment algorithm for stratification can be constructed by selecting
all clauses, all predicates of which are fully determined, and looking for
negative cycles in the dependency graph constructed based on those clauses using
an algorithm such as Bellman-Ford.

@@

Along with constraints, variables, and their domains, two more design decisions
are needed to complete the model: heuristics and restarts. By trial and error,
the variable ordering heuristic was devised to eliminate sources of
\emph{thrashing}, i.e., situations where a contradiction is being `fixed' by
making changes that have no hope of fixing the contradiction. Thus, we partition
all decision variables into an ordered list of groups and require the values of
all variables from one group to be determined before moving to the next group.
Within each group, we use the `fail first' variable ordering heuristic. The
first group consists of all head predicates. Afterwards, we handle all remaining
decision variables from the first clause before proceeding to the next. The
decision variables within each clause are divided into
\begin{enumerate*}[(a)]
\item the $\variable{structure}$ array,
\item body predicates,
\item head arguments,
\item (if $|\variables{}| > 1$) the $\variable{intros}$ array,
\item body arguments.
\end{enumerate*}
For instance, in the clause from \cref{example:sibling}, all visible parts of
the clause would be decided in this order:
\[
  \overset{1}{\mathsf{sibling}}(\overset{3}{X}, \overset{3}{Y}) \gets
  \overset{2}{\mathsf{parent}}(\overset{4}{X}, \overset{4}{Z})
  \overset{2}{\land} \overset{2}{\mathsf{parent}}(\overset{4}{Y},
  \overset{4}{Z}).
\]
We also employ a geometric restart policy, restarting after $10, 10 \times 1.1,
10 \times 1.1^2, \dots$ contradictions.\footnote{Restarts help overcome early
  mistakes in the search process but can be disabled if one wants to find all
  solutions, in which case search is complete regardless of the variable
  ordering heuristic.} We ran close to \num{400000} experiments, investigating
whether the model is efficient enough to generate reasonably-sized programs and
gaining insight into what parameter values make the CSP harder. For
$|\predicates{}|$, $|\variables{}|$, $|\constants{}|$, $\maxNumNodes{}$, and
$\maxNumClauses{} - |\predicates{}|$ (i.e., the number of clauses in addition to
the mandatory $|\predicates{}|$ clauses), we assign all combinations of 1, 2, 4,
8. $\maxArity{}$ is assigned to values 1--4. For each $|\predicates{}|$, we also
iterate over all possible numbers of independent pairs of predicates, ranging
from 0 up to $\binom{|\predicates{}|}{2}$. For each combination of the
above-mentioned parameters, we pick ten random ways to assign arities to
predicates (such that $\maxArity{}$ occurs at least once) and ten random
combinations of independent pairs.

@@

For this experiment, we consider clauses of two types: \emph{rules} are clauses
such that the head atom has at least one variable, and \emph{facts} are clauses
with empty bodies and no variables. We use our constraint model to generate the
rules according to the following parameter values: $|\predicates{}|,
|\variables{}|, \maxNumNodes{} \in \{ 2, 4, 8 \}$, $\maxArity{} \in \{ 1, 2, 3
\}$, $\maxNumClauses{} = |\predicates{}|$, $\constants{} = \emptyset$. Just like
before, we explore all possible numbers of independent predicate pairs. We also
add a constraint that forbids empty bodies. For both rules and facts,
probabilities are uniformly sampled from $\{ 0.1, 0.2, \dots, 0.9 \}$.
Furthermore, all rules are probabilistic, while we vary the proportion of
probabilistic facts among \SI{25}{\percent}, \SI{50}{\percent}, and
\SI{75}{\percent}. For generating facts, we consider $|\constants{}| \in \{100,
200, 400 \}$ and vary the number of facts among $10^3$, $10^4$, and $10^5$ but
with one exception: the number of facts is not allowed to exceed
\SI{75}{\percent} of all possible facts with the given values of
$\predicates{}$, $\arities{}$, and $\constants{}$. Facts are generated using a
simple procedure that randomly selects a predicate, combines it with the right
number of constants, and checks whether the generated atom is already included
or not. We randomly select configurations from the description above and
generate ten programs with a complete restart of the constraint solver before
the generation of each program, including choosing different arities and
independent pairs. Finally, we set the query of each program to a random fact
not explicitly included in the program and consider six natively supported
algorithms and knowledge compilation techniques: binary decision diagrams (BDDs)
\cite{DBLP:journals/tc/Bryant86}, negation normal form (NNF), deterministic
decomposable NNF (d-DNNF) \cite{DBLP:journals/jair/DarwicheM02}, K-Best
\cite{DBLP:conf/pkdd/GutmannKKR08}, and two encodings based on sentential
decision diagrams \cite{DBLP:conf/ijcai/Darwiche11}, one of which encodes the
entire program (SDDX), while the other one encodes only the part of the program
relevant to the query (SDD).\footnote{Forward SDDs (FSDDs) and forward BDDs
  (FBDDs)
  \cite{DBLP:journals/corr/abs-1911-07750,DBLP:conf/ijcai/VlasselaerBKMR15} are
  omitted because the former uses too much memory and the implementation of the
  latter seems to be broken at the time of writing.}

@@

% We checked that our model agrees with the derived combinatorial formula in up
% to a thousand different scenarios. The details of this empirical investigation
% are omitted as they are not crucial to the thrust of this paper.
